---
title: "Boston_Housing_Analysis"
author: "Sangjun Lim"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Packages
Load the necessary packages
```{r message=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(readr)
library(forcats)
library(knitr)
library(car)
```

### The Data
Load the data   
```{r, message = F}
boston <- read_csv("BostonHousing.csv")
```


```{r}
# Square root transformation to handle skewness
boston <- boston %>%
  mutate(CRIM_SQRT = sqrt(CRIM))


# Data cleaning
# Number of missing values from each column
print(colSums(is.na(boston)))

# If there exists missing values, replace it with the mean of the variable
boston <- boston %>%
  mutate(across(everything(), ~ ifelse(is.na(.), mean(., na.rm = TRUE), .)))

# Check if missing values are adjusted
print(colSums(is.na(boston)))

# Checking if there are duplicated rows
print(nrow(boston[duplicated(boston), ]))

# Remove duplicated rows
boston <- boston %>% 
  distinct()

# Check if duplicated rows are successfully removed
print(nrow(boston[duplicated(boston), ]))
```

### Descriptive Statistics and Correlation Matrix
```{r, fig.align='center'}
# Create correlation matrix
cor_matrix <- cor(boston)

# Function to mask the upper triangle and replace NA with space
lower_triangle_with_space <- function(matrix) {
  matrix[upper.tri(matrix)] <- NA
  return(matrix)
}

# Apply the function to the correlation matrix
cor_matrix_lower <- lower_triangle_with_space(cor_matrix)

# Format the correlation matrix to 3 digits
cor_matrix_lower <- format(round(cor_matrix_lower, 3), nsmall = 3)

# Calculate mean and standard deviation for each column
means <- colMeans(boston)
sds <- apply(boston, 2, sd)

# Combine mean, standard deviation and correlation matrix (table format)
result_table <- cbind(
  Mean = format(round(means, 3), nsmall = 3),
  SD = format(round(sds, 3), nsmall = 3),
  cor_matrix_lower
)

# Print the table
kable(result_table, caption = "Descriptive Statistics and Correlation Matrix")
```

### Multiple Linear Regression Model
```{r, fig.align='center'}
# Reshape the data to a long format
boston_long <- boston %>%
  pivot_longer(cols = everything(), names_to = "variable", values_to = "value")

# Histogram for each numerical variable in the dataset
ggplot(boston_long, aes(x = value)) +
  geom_histogram(bins = 10, fill = "skyblue", color = "black") +
  facet_wrap(~ variable, scales = "free", ncol = 4) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(title = "Histograms of Numerical Variables in Boston Housing Dataset",
       x = "Value",
       y = "Frequency")


boston_long <- boston %>%
  pivot_longer(cols = -MEDV, names_to = "variable", values_to = "value")

ggplot(boston_long, aes(x = value, y = MEDV)) +
  geom_point(alpha = 0.8, size = 0.75) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  facet_wrap(~ variable, scales = "free", ncol = 4) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(title = "Scatterplots of MEDV and each Independent Variable",
       x = "Value",
       y = "MEDV")


# Create multiple linear regression model
mlr_model <- lm(MEDV ~ ., data = boston)


```

### Model Diagnostics
```{r}
par(mfrow = c(2, 2))
plot(mlr_model)
```

The plots test the following assumptions:

  1. **Residuals vs. Fitted**: Deviations from a horizontal line indicate non-linearity, suggesting the need for a different modeling approach.

  2. **Normal Q-Q Plot**: Deviations from the diagonal line indicate non-normality of the residuals, which may affect inference.

  3. **Scale-Location Plot**: A non-horizontal line or clustered points suggest heteroscedasticity, indicating that the variance of residuals is not constant.

```{r}
vif <- vif(mlr_model)
print(vif)
```
The VIF (Variance Inflation Factor) values indicate significant multicollinearity among the predictors. High VIF values suggest that some variables are highly correlated, which can compromise the stability of the regression coefficients. Notably, the VIF values for ***CRIM* (15.12)**, ***CRIM_SQRT* (37.68)**, ***RAD* (12.66)**, and ***TAX* (9.00)** are particularly concerning.

```{r}
summary(mlr_model)
```

The model demonstrates a good fit, explaining approximately 73.37% of the variance in the variable *MEDV* with an adjusted R-squared of 0.7267. This indicates that the predictors included in the model are effective in explaining the variability in home prices in Boston.

The coefficients in the model exhibit varying levels of statistical significance. Key predictors like the *RM*, *NOX*, and *LSTAT* are highly significant, with p-values less than 0.001, indicating strong evidence that these variables meaningfully influence variable *MEDV*.

However, the model faces violations of assumptions related to non-linearity, non-normality of residuals, and multicollinearity, which should be addressed to enhance the reliability of the results. These issues may warrant consideration of alternative modeling methodologies to better capture the underlying relationships and improve overall model performance.

